The shape of kelp: environmental influences on kelp morphology

### 2018-11-08 chapter_2.Rmd
addition to intro section

Kelps are sessile species and have to adapt to local environmental conditions in order to persist and survive. The resilience of kelps is evident through their distribution, which spans a range of environmental conditions that vary spatially and temporally. Wave exposure and temperature are regarded as important environmental drivers with regards to macroalgae, and play a role in the growth, reproduction, and survival of kelp species. Wave exposure is recognised as the main cause of kelp mortality, which may manifest itself through dislodgment of individuals or breakage of important morphological structures. These mechanisms usually occur during times of high wave energy or pulse disturbance events, such as storms. Kelps are able to reduce the probability of dislodgment or breakage through morphological adaptation, which may involve several strategies. Kelps can either reduce drag, increase attachment strength or increase flexibility in order to survive high wave energy environments.

The biogeographic distribution of kelp is limited by seawater temperature [@Bolton2010], where increasing temperature gradients reduce kelp distribution. Due to this limiting factor, the two main species of kelps in southern African waters, _Ecklonia maxima_ and _Laminaria pallida_, are distributed along a section of the south coast from De Hoop, extending west around the Cape Peninsula, and thriving north into Namibia [@Molloy1996; @Stegenga1997]. This distribution follows a temperate gradient, where sea temperatures increase as one moves south from Namibia, around Cape Point and towards De Hoop. Although the two species occur together for the majority of the coast, their basic morphologies and resource needs vary to a degree. The larger species, _Ecklonia maxima_, is distributed from Lüderitz to Cape Agulhas [@Bolton1985a; @Probyn1985; @Bolton1987; @Bolton2012]. Characterised by a large distal swollen bulb filled with gas, and smooth fronds, this species grows to approximately 10 meters [@Bolton1987]. There was, however, a 17-meter specimen collected in 2015 off Cape Point (Smit, unpubl. data).This species of kelp not only dominate the biomass of the South African nearshore, but plays an important ecological role [@Bustamante1996]. The estimated productivity of _Ecklonia maxima_ within South Africa varies between 350 and 1500g Cm-2yr-1 [@Mann1982]. Across the majority of the coastline, _Laminaria pallida_ remains a subsurface kelp, dominating the kelp biomass at depths greater than 10 meters [@Field1980b; @Bolton1987; @Molloy1996]. This species is distributed from Danger Point, east of the Cape Peninsula, to Rocky Point in northern Namibia, and reaches depths of greater than 20 meters [@Field1980b; @Molloy1996; @Molloy1996; @Stegenga1997]. Towards the north along the west coast, from around Hondeklipbaai, _Laminaria pallida_ replaces _Ecklonia maxima_ as the dominant kelp species [@Velimirov1977; @Stegenga1997] and it also occupies increasingly shallower subtidal regions. The northern populations also exhibit an increase in stipe hollowness, compared to the solid stipe morphs in the species’ southern distributions [@Molloy1996]. This variation in morphology was thought to represent two distinct species, with the northern populations formerly described as _Laminaria schinzii_ Foslie [@Molloy1996]. Genetic work has subsequently shown that the two morphs are in fact the same species [@Rothman2017]. In southern African waters, the primary production of _Laminaria pallida_ is between 120 and 1900g C m2yr1, similar to that of _Ecklonia maxima_ [@Mann1982]. Primary production is not the only pathway.

@@

@@

In Figure 1 the total and seasonal wave roses of the directional wave buoy just offshore from Cape Town is given form the years 2000 till end 2017 (the period for which directional wave spectra was available). It is due to the near consistent South-westerly swell waves and the complex orography around the peninsula, that the wave energy distribution around the Cape Peninsula varies significantly over a small geographical area. For details on the seasonality of the wave climate around the Cape peninsula please refer to [@Veitch2018].

@@

In Figure 2 the total coastal wave exposure of the Cape Peninsula is given in terms of wave energy (kW per meter wave crest length). These values have been computed using model output from approximately 20 year of simulated nearshore wave conditions. Hindcast global wave parameters were refracted into the nearshore via a spectral phase-averaged numerical wave modelling code SWAN (Simulating, WAves in the Nearshore - 3rd generation). The model is fully spectral in all frequencies and directions (0o -360o) (can solve swell and local wind generated waves simultaneously, propagating in different direction) and was used to model the propagation of short crested swell waves, waves generated by wind, non-linear wave-wave interaction and dissipation and depth induced breaking and refraction. Wave energy dissipation due to bottom friction and whitecapping is also accounted for [@Deltares2014].

@@

The directional sheltering effect of the Cape Peninsula, against the dominant swell direction, (given in Figure 1) is clear observed in the wave exposures presented in Figure 2. The classification from fully sheltered to extremely exposed is based on the total wave energy upper and lower limits. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [@Norderhaug2012; @Liberti2013; @Sundblad2014]. Kelp and other ecosystems adapt to their typical environmental exposure and thus what is classified as extreme abiotic coastal forcings (temperature, winds, waves and currents) will vary around the world depending on the geographical setting.

>>>>

Kelps are sessile species and have to adapt to local environmental conditions in order to persist and survive. The resilience of kelps is evident through their distribution, which spans a range of environmental conditions that vary spatially and temporally. Wave exposure and temperature are regarded as important environmental drivers with regards to macroalgae, and play a role in the distribution, abundance, diversity, composition, growth [@Cousens1982] and productivity [@pedersen2012a] of kelps. Wave exposure is recognised as the main cause of kelp mortality, which may manifest itself through dislodgment of individuals or breakage of important morphological structures. These mechanisms usually occur during times of high wave energy or pulse disturbance events, such as storms.Mechanical forces generated by the hydrodynamic environment, in the form of sudden strong ocean currents or storms, between 10- 20 m s-1 with accelerations of 400 m s-2 [@Friedland1995] are the biggest threat to kelp survival. 
Kelps are able to reduce the probability of dislodgment or breakage through morphological adaptation, which may involve several strategies. Kelps can either reduce drag, increase attachment strength or increase flexibility in order to survive high wave energy environments.The strategy that the algal organism employs is dependent on the species. For instance, branched, articulated algae will vary its morphology differently to that of foliose, more flexible macroalgae. For example, @moss1948 investigated the anatomy, chemical composition of *Fucus spiralis* at three sites that varied in wave exposure (sheltered, medium exposure and exposed). The authors found that individuals in exposed sites showed less branching of thalli as well as variation physiological components, such as organic nitrogen, mannitol, laminarian and alginic acid concentrations. The authors also noted a ‘crumpling effect’ displayed by individuals from exposed sites and inferred that this strategy may reduce overall drag.  Other studies show that macroalgae in wave exposed environments have morphologies that reduce overall drag, increase strength of attachment or increase flexibility. For example, a study by @FowlerWalker2006 tested for differences in morphology of *Ecklonia radiata* between wave-sheltered and wave-exposed sites and through a combination of *in situ* sampling and transplantation of juvenile plants. The results showed that morphology differed between wave-sheltered and wave-exposed sites (thin thallus at sheltered sites and a narrow, thick thallus with a thick stipe at exposed sites), and was consistent with previous studies. Juveniles transplanted into wave exposed sites under went rapid morphological adaptation, whilst the opposite was true for wave-sheltered sites which showed slower morphological adaptation. Dislodgement through increased wave exposure is not the only hydrological parameter playing a role in kelp mortality. A study by @DeBettignies2013 tested if there is a linked between orbital water velocity representative of wave-swept reefs and the morphology and biomass of *Ecklonia radiata*. The results showed that morphological adaptation through drag reduction and strength increase strategies do not reduce drag as efficiently as reduction in overall biomass through reduction in thallus surface area. The authors concluded that at peak orbitat water velocities size (biomass) and not morphological adaptation was the main driver of hydrological performance in wave-swept environments. There is also evidence that morphological adaptation is driven by currents, and in fact may be driving hydrological performance of macrolgae.

### The South African context

The biogeographic distribution of kelp is limited by seawater temperature [@Bolton2010], where increasing temperature gradients reduce kelp distribution. Due to this limiting factor, the two main species of kelps in southern African waters, _Ecklonia maxima_ and _Laminaria pallida_, are distributed along a section of the south coast from De Hoop, extending west around the Cape Peninsula, and thriving north into Namibia [@Molloy1996; @Stegenga1997]. This distribution follows a temperate gradient, where sea temperatures increase as one moves south from Namibia, around Cape Point and towards De Hoop. Although the two species occur together for the majority of the coast, their basic morphologies and resource needs vary to a degree. The larger species, _Ecklonia maxima_, is distributed from Lüderitz to Cape Agulhas (Fig. 1) [@Bolton1985a; @Probyn1985; @Bolton1987; @Bolton2012].

@@

Characterised by a large distal swollen bulb filled with gas, and smooth fronds, this species grows to approximately 10 meters [@Bolton1987]. There was, however, a 17-meter specimen collected in 2015 off Cape Point (Smit, unpubl. data).This species of kelp not only dominate the biomass of the South African nearshore, but plays an important ecological role [@Bustamante1996]. The estimated productivity of _Ecklonia maxima_ within South Africa varies between 350 and 1500g Cm-2yr-1 [@Mann1982]. Across the majority of the coastline, _Laminaria pallida_ remains a subsurface kelp, dominating the kelp biomass at depths greater than 10 meters [@Field1980b; @Bolton1987; @Molloy1996]. This species is distributed from Danger Point, east of the Cape Peninsula, to Rocky Point in northern Namibia, and reaches depths of greater than 20 meters [@Field1980b; @Molloy1996; @Molloy1996; @Stegenga1997]. Towards the north along the west coast, from around Hondeklipbaai, _Laminaria pallida_ replaces _Ecklonia maxima_ as the dominant kelp species [@Velimirov1977; @Stegenga1997] and it also occupies increasingly shallower subtidal regions. The northern populations also exhibit an increase in stipe hollowness, compared to the solid stipe morphs in the species’ southern distributions [@Molloy1996]. This variation in morphology was thought to represent two distinct species, with the northern populations formerly described as _Laminaria schinzii_ Foslie [@Molloy1996]. Genetic work has subsequently shown that the two morphs are in fact the same species [@Rothman2017]. In southern African waters, the primary production of _Laminaria pallida_ is between 120 and 1900g C m2yr1, similar to that of _Ecklonia maxima_ [@Mann1982]. Primary production is not the only pathway.

@@

In Figure 2 the total and seasonal wave roses of the directional wave buoy just offshore from Cape Town is given form the years 2000 till end 2017 (the period for which directional wave spectra was available). It is due to the near consistent South-westerly swell waves and the complex orography around the peninsula, that the wave energy distribution around the Cape Peninsula varies significantly over a small geographical area. For details on the seasonality of the wave climate around the Cape peninsula please refer to [@Veitch2018].

@@

In Figure 3 the total coastal wave exposure of the Cape Peninsula is given in terms of wave energy (kW per meter wave crest length). These values have been computed using model output from approximately 20 year of simulated nearshore wave conditions. Hindcast global wave parameters were refracted into the nearshore via a spectral phase-averaged numerical wave modelling code SWAN (Simulating, WAves in the Nearshore - 3rd generation). The model is fully spectral in all frequencies and directions (0o -360o) (can solve swell and local wind generated waves simultaneously, propagating in different direction) and was used to model the propagation of short crested swell waves, waves generated by wind, non-linear wave-wave interaction and dissipation and depth induced breaking and refraction. Wave energy dissipation due to bottom friction and whitecapping is also accounted for [@Deltares2014].

@@

The directional sheltering effect of the Cape Peninsula, against the dominant swell direction, (given in Figure 2) is clear observed in the wave exposures presented in Figure 2. The classification from fully sheltered to extremely exposed is based on the total wave energy upper and lower limits. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [@Norderhaug2012; @Liberti2013; @Sundblad2014]. Kelp and other ecosystems adapt to their typical environmental exposure and thus what is classified as extreme abiotic coastal forcings (temperature, winds, waves and currents) will vary around the world depending on the geographical setting.
